---
title: "Generating Phenotype Template Cohorts on Eunomia"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating Phenotype Template Cohorts on Eunomia}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("Eunomia", quietly = TRUE) &&
    requireNamespace("CohortGenerator", quietly = TRUE) &&
    requireNamespace("DatabaseConnector", quietly = TRUE) &&
    requireNamespace("Capr", quietly = TRUE)
)
```

This vignette demonstrates both cohort-building entry points in this package
against the [Eunomia](https://github.com/OHDSI/Eunomia) synthetic CDM, using a
small set of **contrived** concept sets for an illustrative "Acute
Gastrointestinal Bleed" phenotype. The concept IDs below are chosen to be
plausible OMOP standard concepts for demonstration purposes only — they are
not a clinically validated phenotype definition.

1. `buildPhenotypeTemplates()` — the evidence-strictness *ladder* crossed with
   index specificity, hospitalization, and alternative-diagnosis exclusion.
2. `buildEvidenceCombinationCohorts()` — the curated *combination/threshold*
   set (base case, single-evidence-only cohorts, and "at least K of 4"
   cohorts).

Both functions return a `CohortGenerator`-compatible cohort definition set,
which we then generate on Eunomia with `CohortGenerator::runCohortGeneration()`.

## Setup

```{r setup}
library(Capr)
library(CohortGenerator)
library(PhenotypeTemplates)
```

Download/connect to the Eunomia sample CDM:

```{r eunomia-connect}
connectionDetails <- Eunomia::getEunomiaConnectionDetails()
cdmDatabaseSchema <- "main"
cohortDatabaseSchema <- "main"
cohortTable <- "phe_tpl_vignette_cohorts"
```

## Contrived concept sets

Each clinical category (see `?buildPhenotypeTemplates`) is a `Capr::cs()`
concept set built from `descendants()` of a single seed concept ID, purely
for illustration:

```{r concept-sets}
gib_I <- cs(descendants(192671L), name = "GI hemorrhage - disease of interest")
gib_H <- cs(descendants(4171594L), name = "GI tract disorder - hypernym")
gib_S <- cs(descendants(200219L), name = "Abdominal pain - symptom")
gib_D <- cs(descendants(4028949L), name = "Endoscopy - diagnostic procedure")
gib_T <- cs(descendants(923645L), name = "Pantoprazole - treatment")
gib_C <- cs(descendants(439777L), name = "Anemia - complication")
gib_A <- cs(descendants(4027663L), name = "Peptic ulcer - alternative diagnosis")
gib_E <- cs(descendants(1118084L), name = "NSAID exposure - etiology")

phenotypeLabel <- "Acute GI Bleed (contrived)"
```

## Use case 1: evidence-strictness ladder (`buildPhenotypeTemplates()`)

`buildPhenotypeTemplates()` crosses an evidence ladder (L0 index-only, L1
treatment/follow-up, L2 symptom/diagnostic + treatment/follow-up) with index
specificity (I vs I|H), hospitalization variants, and alternative-diagnosis
exclusion:

```{r ladder-config}
ladderConfig <- phenotypeConfig(
  clinicalCourse = "transient_single",
  expectedCareSetting = "acute_care_common",
  fixedExitDays = 30,
  hasDescreteRecordedSymptoms = TRUE,
  requiresDiagnosticTestOrProcedure = TRUE,
  requiresActiveTreatmentWithin30d = TRUE,
  expectsConditionSpecificFollowupOrSequelae1yr = TRUE
)

ladderCohorts <- buildPhenotypeTemplates(
  cs_I = gib_I, cs_H = gib_H, cs_S = gib_S, cs_D = gib_D,
  cs_T = gib_T, cs_C = gib_C, cs_A = gib_A, cs_E = gib_E,
  phenotypeLabel = phenotypeLabel,
  config = ladderConfig,
  startCohortId = 0
)

ladderCohorts[, c("cohortId", "cohortName")]
```

## Use case 2: curated evidence combinations (`buildEvidenceCombinationCohorts()`)

`buildEvidenceCombinationCohorts()` instead builds a base case, four
single-evidence-only cohorts, and a set of "at least K of 4 evidence
components" threshold cohorts, using `Capr::withAtLeast()` under the hood:

```{r combo-cohorts}
comboCohorts <- buildEvidenceCombinationCohorts(
  cs_I = gib_I, cs_H = gib_H, cs_S = gib_S, cs_D = gib_D,
  cs_T = gib_T, cs_C = gib_C, cs_A = gib_A, cs_E = gib_E,
  phenotypeLabel = phenotypeLabel,
  config = ladderConfig,
  startCohortId = max(ladderCohorts$cohortId)
)

comboCohorts[, c("cohortId", "cohortName")]
```

## Generate both cohort sets on Eunomia

Combine the two cohort definition sets and generate them together with
`CohortGenerator`:

```{r generate}
allCohorts <- rbind(ladderCohorts, comboCohorts)

cohortTableNames <- CohortGenerator::getCohortTableNames(cohortTable = cohortTable)
CohortGenerator::createCohortTables(
  connectionDetails = connectionDetails,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTableNames = cohortTableNames
)

CohortGenerator::runCohortGeneration(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = cdmDatabaseSchema,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTableNames = cohortTableNames,
  cohortDefinitionSet = allCohorts,
  incremental = FALSE,
  outputFolder = "./eunomia_pval_templates"
)

CohortGenerator::getCohortCounts(
  connectionDetails = connectionDetails,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTable = cohortTableNames$cohortTable,
  cohortDefinitionSet = allCohorts
)
```
