---
title: "Building Individual Cohorts with outcomePhenotypeTpl()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Individual Cohorts with outcomePhenotypeTpl()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("Eunomia", quietly = TRUE) &&
    requireNamespace("CohortGenerator", quietly = TRUE) &&
    requireNamespace("DatabaseConnector", quietly = TRUE) &&
    requireNamespace("Capr", quietly = TRUE)
)
```

The other vignette (`vignette("evidence-combination-cohorts")`) covers the two
batch cohort-set builders, `buildPhenotypeTemplates()` and
`buildEvidenceCombinationCohorts()`. This vignette instead walks through the
standard building blocks in `R/CaprFunctions.R` that those batch builders sit
on top of, so you can construct one-off cohort variants by hand with
`outcomePhenotypeTpl()`.

We use a small set of **contrived** concept sets for an illustrative "Atrial
Fibrillation" phenotype (the categories mirror the `I`/`H`/`S`/`D`/`T`/`C`/`A`/`E`
scheme documented at the top of `R/CaprFunctions.R`). The concept IDs are
plausible OMOP standard concepts chosen for demonstration only — not a
clinically validated phenotype definition.

## Setup

```{r setup}
library(Capr)
library(CohortGenerator)
library(PhenotypeTemplates)
```

```{r eunomia-connect}
connectionDetails <- Eunomia::getEunomiaConnectionDetails()
cdmDatabaseSchema <- "main"
cohortDatabaseSchema <- "main"
cohortTable <- "phe_tpl_basics_vignette_cohorts"
```

## Contrived concept sets

```{r concept-sets}
afib_I <- cs(descendants(313217L), name = "Atrial fibrillation - disease of interest")
afib_H <- cs(descendants(4232697L), name = "Cardiac arrhythmia - hypernym")
afib_S <- cs(descendants(313418L), name = "Palpitations - symptom")
afib_D <- cs(descendants(4013514L), name = "Electrocardiogram - diagnostic procedure")
afib_T <- cs(descendants(1310149L), name = "Warfarin - treatment")
afib_C <- cs(descendants(372924L), name = "Stroke - complication")
afib_A <- cs(descendants(442077L), name = "Panic disorder - alternative diagnosis")
afib_E <- cs(descendants(4142479L), name = "Hyperthyroidism - etiology")

phenotypeLabel <- "Atrial Fibrillation (contrived)"
```

## Step 1: resolve concept set overlaps

Categories share the condition/observation domain and can legitimately
overlap (e.g. a code used for both `I` and `H`). `resolveConceptSetOverlaps()`
applies the fixed precedence rules (I > A > H > E > {S, C}) so the same
concept never counts toward two categories at once:

```{r resolve-overlaps}
resolved <- resolveConceptSetOverlaps(
  cs_I = afib_I, cs_H = afib_H, cs_S = afib_S, cs_D = afib_D,
  cs_T = afib_T, cs_C = afib_C, cs_A = afib_A, cs_E = afib_E
)
```

## Step 2: resolve exit strategy and demographics

`resolvePhenotypeExitStrategy()` turns a `clinicalCourse` (plus optional
overrides) into an era-collapse gap and a Capr end strategy.
`greateDemographicCriteria()` builds the optional age/gender entry criteria:

```{r exit-and-demographics}
exitConfig <- resolvePhenotypeExitStrategy(
  clinicalCourse = "persistent_transient",
  minimumInterepisodeDayGap = 180
)

demographicCriteria <- greateDemographicCriteria(minAge = 18, male = FALSE, female = FALSE)
```

## Step 3: build a single cohort variant with outcomePhenotypeTpl()

`outcomePhenotypeTpl()` is the standard template used by both batch builders.
Called directly, it produces one Capr cohort for exactly the evidence flags
you pass in — here, a hospitalized, symptom/diagnostic + treatment/follow-up
variant with alternative-diagnosis exclusion:

```{r single-cohort}
afibStrict <- outcomePhenotypeTpl(
  cs_I = resolved$cs_I, cs_H = resolved$cs_H, cs_S = resolved$cs_S,
  cs_D = resolved$cs_D, cs_T = resolved$cs_T, cs_C = resolved$cs_C,
  cs_A = resolved$cs_A, cs_E = resolved$cs_E,
  phenotypeLabel = phenotypeLabel,
  templateSuffix = "strict-hosp-Aexcl",
  useHypernymAsIndex = FALSE,
  hasDescreteRecordedSymptoms = TRUE,
  requiresDiagnosticTestOrProcedure = TRUE,
  requiresActiveTreatmentWithin30d = TRUE,
  expectsConditionSpecificFollowupOrSequelae1yr = TRUE,
  treatmentFollowUpOp = "all",
  excludeA = TRUE,
  requiresHospitalization = TRUE,
  demographicCriteria = demographicCriteria,
  eraDays = exitConfig$eraDays,
  endStrategy = exitConfig$endStrategy
)
```

A second, looser variant — index on `I` or `H`, no hospitalization
requirement, "at least 2 of 4" evidence components (via `evidenceThreshold`,
see the combination vignette) instead of the strict AND/OR rule above:

```{r second-cohort}
afibLoose <- outcomePhenotypeTpl(
  cs_I = resolved$cs_I, cs_H = resolved$cs_H, cs_S = resolved$cs_S,
  cs_D = resolved$cs_D, cs_T = resolved$cs_T, cs_C = resolved$cs_C,
  cs_A = resolved$cs_A, cs_E = resolved$cs_E,
  phenotypeLabel = phenotypeLabel,
  templateSuffix = "loose-IH-atLeast2",
  useHypernymAsIndex = TRUE,
  hasDescreteRecordedSymptoms = TRUE,
  requiresDiagnosticTestOrProcedure = TRUE,
  requiresActiveTreatmentWithin30d = TRUE,
  expectsConditionSpecificFollowupOrSequelae1yr = TRUE,
  evidenceThreshold = 2,
  excludeA = FALSE,
  requiresHospitalization = FALSE,
  demographicCriteria = demographicCriteria,
  eraDays = exitConfig$eraDays,
  endStrategy = exitConfig$endStrategy
)
```

## Step 4: assemble a cohort definition set and generate on Eunomia

Each object returned by `outcomePhenotypeTpl()` is a plain Capr cohort with a
`cohortName` attribute; convert them to the `CohortGenerator` cohort
definition set shape and generate:

```{r assemble-and-generate}
toCohortRow <- function(tpl, cohortId) {
  data.frame(
    cohortId = cohortId,
    cohortName = attr(tpl, "cohortName"),
    json = Capr::toCohortJson(tpl),
    sql = CirceR::buildCohortQuery(
      expression = Capr::toCohortJson(tpl),
      options = CirceR::createGenerateOptions(generateStats = FALSE)
    )
  )
}

manualCohorts <- rbind(
  toCohortRow(afibStrict, 1L),
  toCohortRow(afibLoose, 2L)
)

cohortTableNames <- CohortGenerator::getCohortTableNames(cohortTable = cohortTable)
CohortGenerator::createCohortTables(
  connectionDetails = connectionDetails,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTableNames = cohortTableNames
)

CohortGenerator::runCohortGeneration(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = cdmDatabaseSchema,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTableNames = cohortTableNames,
  cohortDefinitionSet = manualCohorts,
  incremental = FALSE,
  outputFolder = "./eunomia_pval_templates"
)

CohortGenerator::getCohortCounts(
  connectionDetails = connectionDetails,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTable = cohortTableNames$cohortTable,
  cohortDefinitionSet = manualCohorts
)
```
